---
title: "Data transformation with dplyr"
author: "Wasilios Hariskos"
date: "8 November 2018"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# The gapminder dataset

```{r echo=TRUE}
# install.packages("gapminder")
library(gapminder)
```

```{r echo=TRUE}
gapminder 
```

In this module we will use data from gapminder, which tracks demographic data in countries of the world over time. 
With `?gapminder` you can find out more about the gapminder dataset.
Notice that the gaminder dataset is stored as a tibble, a modern take on data frames. Use `?tibble` to find out more about tibbles. 

# The dplyr package

```{r echo=TRUE, message=FALSE, warning=FALSE}
# install.packages("dplyr")
library(dplyr)
```

## Get a glimpse of you data
```{r echo=TRUE}
glimpse(gapminder)
```

## Comparison: glimpse() vs str()

```{r echo=TRUE}
str(gapminder)
```

Try `?glimpse` to find out more.

# Filtering observations with filter()

## One criterion: Country equals Germany

```{r echo=TRUE, message=FALSE, warning=FALSE}
gapminder %>%
  filter(country == "Germany")
```

## Two criteria: Country equals Germany and year >= 1982

```{r echo=TRUE}
gapminder %>%
  filter(country == "Germany", year >= 1982)
```

# Sorting obsevations with arrange()

## Sorting from low to high life expectancy

```{r echo=TRUE}
gapminder %>%
  filter(year == 1997, continent == "Europe") %>% 
  arrange(lifeExp) %>%
  print(n = 20)
```

## Sorting from high to low life expectancy 

```{r echo=TRUE}
countries <- c("Germany", "Greece", "Portugal", "Spain", "Singapore", "Serbia", "India", "United States")

gapminder %>%
  filter(year == 1982, country %in% countries) %>% 
  arrange(desc(lifeExp))
```

# Change or create a variable with mutate()

```{r echo=TRUE}
gapminder %>%
  mutate(pop = pop / 1000000)
```


```{r echo=TRUE}
gapminder %>%
  mutate(popMM = pop / 1000000)
```

# Combining filter(), arrange(), and mutate()

```{r echo=TRUE}
gdp_1997 <- gapminder %>%
  mutate(gdp = gdpPercap * pop) %>%
  filter(year == 1997) %>%
  arrange(desc(gdp))
gdp_1997
```

```{r echo=TRUE}
# Excurse: Scientific form (exponential form) 
non_scientific <- format(gdp_1997$gdp[1:3], scientific = FALSE)
non_scientific

scientific <- c(9.7613, 3.6296, 2.8159) * 10^12
scientific
```

# Use summarize() to calculate statistics

## Average life expectancy accross all countries in 1997

```{r echo=TRUE}
gdp_1997 %>%
  summarize(meanLifeExp = mean(lifeExp))
```
## Add average total population accross all countries in 1997

```{r echo=TRUE}
gdp_1997 %>%
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(pop))
```

## Add more statistics

```{r echo=TRUE}
gdp_1997 %>%
  summarize(meanLifeExp = mean(lifeExp),
            medianLifeExp = median(lifeExp),
            minGdpPercap = min(gdpPercap),
            maxGdpT = max(gdp/1e12),
            totalPopB = sum(pop/1e9)) 
```
Note: Show `NA` bug (live coding) if `totalPopMM = sum(pop)/1e6` or `totalPop = sum(pop)`. Show quick and dirty solution with `totalPop = sum(gdp_1997$pop)`. 

## Excourse: Abbreviations for large numbers

+ Thousand: 1,000 or 1e03 or 1K or 1M. M being the Roman numeral for 1,000.
+ Million: 1,000,000 or 1e06 or 1MM. MM means "thousands (times) thousands".
+ Billion: 1,000,000,000 or 1e09 or 1B.
+ Trillion: 1,000,000,000,000 or 1e12 or 1T.

## Summarize by year using group_by()


```{r echo=TRUE}
gapminder %>%
  group_by(year) %>%
  summarize(meanLifeExp = mean(lifeExp),
            medianLifeExp = median(lifeExp),
            minGdpPercap = min(gdpPercap),
            maxGdpT = max((gdpPercap * pop)/1e12),
            totalPopB = sum(pop/1e9)) 
```

## Summarize by continent and year using group_by()

```{r echo=TRUE}
by_continent_year <- gapminder %>%
  group_by(continent, year) %>%
  summarize(meanLifeExp = mean(lifeExp),
            medianLifeExp = median(lifeExp),
            minGdpPercap = min(gdpPercap),
            maxGdpT = max((gdpPercap * pop)/1e12),
            totalPopB = sum(pop/1e9))

print(by_continent_year, n = 20)
```

# Outlook: Visualizing summarized data with ggplot2
```{r echo=TRUE, message=FALSE, warning=FALSE}
# install.packages("ggplot2")
library(ggplot2)
```

Initialize ggplot object and declare the input data frame. 
Try `?ggplot` for more information. 

```{r echo=TRUE}
ggplot(data = by_continent_year)
```

Add point geom to create a scatterplot with aesthetics mappings x-axis equals year and y-axis equals mean life expectancy. 
Try `?geom_point` and try `?aes` for more information.

```{r echo=TRUE}
ggplot(data = by_continent_year) +
  geom_point(mapping = aes(x = year, 
                           y = meanLifeExp))
```

Map continent to color aesthetic.

```{r echo=TRUE}
ggplot(data = by_continent_year) +
  geom_point(mapping = aes(x = year, 
                           y = meanLifeExp, 
                           color = continent))
```

Map total population in billion to size aesthetic.

```{r echo=TRUE}
ggplot(data = by_continent_year) +
  geom_point(mapping = aes(x = year, 
                           y = meanLifeExp, 
                           color = continent,
                           size = totalPopB))
```